---
title: "Gather Across Tiers and Quintiles"
author: "Matt Lawrence"
date: "3/7/2019"
output: pdf_document
---

Start with a dataframe that has the count - for each tier - of kids from each parent quintile ending up in the top 1 percent. This is going to work the cleanest if you only have these variables in your dataframe.

```{r load data and packages}
library(tidyverse)
tier1pc <- read.csv("chris_gather.csv")
```

The trick is to turn the wide data into long data using gather. In this next chunk, you create a new data frame called `tier_gather` based on the existing `tier1pc` data frame. The gather line says:

- Create a new variable called `par_quintile`
- Create a new variable called `count`
- Each of the variable names from `tier1pc` between count_ktop1_cond_parq1 and count_ktop1_cond_parq5 will become the values of the new variable called `par_quintile`
- Each of the values associated with those variable names in `tier1pc` will become the values of the new `count` variable for the observations where the value of `par_quintile` is the original variable 

```{r gather}
tier_gather <- tier1pc %>%
     gather(par_quintile, count, count_ktop1_cond_par_q1:count_ktop1_cond_par_q5)
```

If it worked correctly, we should now have 25 observations. That is one observation for each parent quintile for each tier.
 
Let's clean up some of the labels before making the plot.
```{r rename variables and levels}
tier_gather <- tier_gather %>%
     mutate(par_quintile = factor(par_quintile,
                                  labels = c("Q1", "Q2", "Q3", "Q4", "Q5")),
            tier = factor(tier,
                          labels = c("Tier A", "Tier B", "Tier C", "Tier D", "Tier E")))
```

And let's make the plot.
```{r plot it}
p1 <- ggplot(tier_gather, aes(x = par_quintile, y = count, fill = tier))
p1 + geom_col()
```

```{r}
my_gather <- tier1pc %>%
  gather(par_quintile, count, count_ktop1_cond_par_q1:count_ktop1_cond_par_q5)
```
```{r p2}
 
```