---
title: "Legible_table3"
author: "Chris"
date: "3/12/2019"
output: pdf_document
---

```{r load data and packages, echo = FALSE, warning = FALSE, message = FALSE}
chettey <- read.csv("mrc_table3.csv")
library(tidyverse)
```

```{r select data}
chettey2 <- chettey %>%
  select(tier, par_q1:par_q5, type, kq5_cond_parq1,kq5_cond_parq2, kq5_cond_parq3,kq5_cond_parq4,kq5_cond_parq5,count)
```

Give collges with insufficient data and no-college types.
***Is there a better way to group these? 
```{r}
chettey2 <- chettey2 %>%
 mutate(type = ifelse(type %in% 1:3, type,
               ifelse(tier==13, 4,
               5)))
```

Table with rows associated with college-type and columns associated with parent quintile. The numbers in each represent the number of kids who make it to the top quintile from associated collge-type and 
```{r}
chettey3 <- chettey2 %>%
  mutate(ct_kq5_cond_parq1 = count * kq5_cond_parq1,
         ct_kq5_cond_parq2 = count * kq5_cond_parq2,
         ct_kq5_cond_parq3 = count * kq5_cond_parq3,
         ct_kq5_cond_parq4 = count * kq5_cond_parq4,
         ct_kq5_cond_parq5 = count * kq5_cond_parq5) %>%
  group_by(type) %>%
  summarize(pq1= sum(ct_kq5_cond_parq1, na.rm=TRUE),
            pq2= sum(ct_kq5_cond_parq2, na.rm=TRUE),
            pq3= sum(ct_kq5_cond_parq3, na.rm=TRUE),
            pq4= sum(ct_kq5_cond_parq4, na.rm=TRUE),
            pq5= sum(ct_kq5_cond_parq5, na.rm=TRUE))
chettey3
```

```{r}
middlebury <- chettey %>%
  filter(name == "Middlebury College", cohort == 1991)

middlebury$par_top1pc
```


Gather the data
```{r}
chettey3 <- chettey3 %>%
  gather(par_quintile, count, pq1:pq5)
```

Plot it
```{r}
ggplot(chettey3, aes(par_quintile, count, fill=type))+geom_col()+
  labs(title= "Distribution of Kids Who End Up in Top Quintile",
       x = "Parent's Quintile", y="Count in Top Quintile")
```

Same data but packaged a little differently
```{r}
ggplot(chettey3, aes(type, count, fill=par_quintile))+geom_col()+
   labs(title= "Distribution of Kids Who End Up in Top Quintile",
       x = "Parent's Quintile", y="Count in Top Quintile")
```

#Where does kids from top quintile end up?
```{r}
chettey4 <- chettey %>%
  select(tier, par_q1:par_q5, type, kq4_cond_parq5,kq3_cond_parq5,kq2_cond_parq5,kq1_cond_parq5,kq5_cond_parq5,count) %>%
  mutate(type = ifelse(type %in% 1:3, type,
               ifelse(tier==13, 4,
               5))) %>%
  mutate(ct_kq1_cond_parq5 = count * kq1_cond_parq5,
         ct_kq2_cond_parq5 = count * kq2_cond_parq5,
         ct_kq3_cond_parq5 = count * kq3_cond_parq5,
         ct_kq4_cond_parq5 = count * kq4_cond_parq5,
         ct_kq5_cond_parq5 = count * kq5_cond_parq5) %>%
  group_by(type) %>%
  summarize(kq1= sum(ct_kq1_cond_parq5, na.rm=TRUE),
            kq2= sum(ct_kq2_cond_parq5, na.rm=TRUE),
            kq3= sum(ct_kq3_cond_parq5, na.rm=TRUE),
            kq4= sum(ct_kq4_cond_parq5, na.rm=TRUE),
            kq5= sum(ct_kq5_cond_parq5, na.rm=TRUE))
pander(chettey4)
```

Gather the data
```{r}
chettey4 <- chettey4 %>%
  gather(k_quintile, count, kq1:kq5)
```

Plot it
```{r}
ggplot(chettey4, aes(type,count,fill=k_quintile))+geom_col()
```

```{r}
ggplot(chettey4, aes(k_quintile,count,fill=type))+geom_col()
```

Potential Research Question: 
How is our current higher education system conducive to the social reproduction of the elite? Is this a problem? If so, what could be done?

Potential Key Words:
Higher Education
Admissions
Parenting (Concerted Cultivation)
Income
Child outcomes
Social capital

Potential Project Sources

The Future of College Access: The Declining Role of Public Higher Education in Promoting Equal Opportunity
https://journals.sagepub.com/doi/abs/10.1177/0002716202238569


Dream Hoarders (might need to find a different version since this is just chapter 1. Maybe ask Matt about where to look)
https://www.brookings.edu/wp-content/uploads/2016/05/chapter-one_-dream-hoarders.pdf


Privilege and domination: The role of the upper class in American higher education
https://journals.sagepub.com/doi/abs/10.1177/053901847501400607


Concerted Cultivation? Parenting Values, Education and Class Diversity
https://journals.sagepub.com/doi/abs/10.1177/0038038511399618


Does Parenting Style Matter? Concerted Cultivation, Educational Expectations, and the Transmission of Educational Advantage
https://journals.sagepub.com/doi/full/10.1177/0731121414562967


Socioeconomic Inequality In Education
http://go.galegroup.com/ps/retrieve.do?tabID=T003&resultListType=RESULT_LIST&searchResultsType=SingleTab&searchType=BasicSearchForm&currentPosition=3&docId=GALE%7CCX3273000122&docType=Topic+overview&sort=RELEVANCE&contentSegment=&prodId=GVRL&contentSet=GALE%7CCX3273000122&searchId=R2&userGroupName=vol_m58c&inPS=true

Inequality of Opportunity in Comparative Perspective: Recent Research on Educational Attainment and SocialMobility
https://www.annualreviews.org/doi/abs/10.1146/annurev.soc.31.041304.122232

Chettey
https://www.nytimes.com/interactive/2017/01/18/upshot/some-colleges-have-more-students-from-the-top-1-percent-than-the-bottom-60.html?action=click&contentCollection=Opinion&module=Trending&version=Full&region=Marginalia&pgtype=article







