---
title: "Significance_Tests"
author: "Chris Gonzales"
date: "4/11/2019"
output: pdf_document
---

```{r load data and packages, echo = FALSE, warning = FALSE, message = FALSE}
chettey <- read.csv("mrc_table3.csv")
library(tidyverse)
```


Add counts to the data
```{r}
chettey2 <- chettey %>%
  mutate(ct_kq5_cond_parq1 = count * kq5_cond_parq1,
         ct_kq5_cond_parq2 = count * kq5_cond_parq2,
         ct_kq5_cond_parq3 = count * kq5_cond_parq3,
         ct_kq5_cond_parq4 = count * kq5_cond_parq4,
         ct_kq5_cond_parq5 = count * kq5_cond_parq5,
         
         ct_kq4_cond_parq1 = count * kq4_cond_parq1,
         ct_kq4_cond_parq2 = count * kq4_cond_parq2,
         ct_kq4_cond_parq3 = count * kq4_cond_parq3,
         ct_kq4_cond_parq4 = count * kq4_cond_parq4,
         ct_kq4_cond_parq5 = count * kq4_cond_parq5,
         
         ct_kq3_cond_parq1 = count * kq3_cond_parq1,
         ct_kq3_cond_parq2 = count * kq3_cond_parq2,
         ct_kq3_cond_parq3 = count * kq3_cond_parq3,
         ct_kq3_cond_parq4 = count * kq3_cond_parq4,
         ct_kq3_cond_parq5 = count * kq3_cond_parq5,
         
         ct_kq2_cond_parq1 = count * kq2_cond_parq1,
         ct_kq2_cond_parq2 = count * kq2_cond_parq2,
         ct_kq2_cond_parq3 = count * kq2_cond_parq3,
         ct_kq2_cond_parq4 = count * kq2_cond_parq4,
         ct_kq2_cond_parq5 = count * kq2_cond_parq5,
         
         ct_kq1_cond_parq1 = count * kq1_cond_parq1,
         ct_kq1_cond_parq2 = count * kq1_cond_parq2,
         ct_kq1_cond_parq3 = count * kq1_cond_parq3,
         ct_kq1_cond_parq4 = count * kq1_cond_parq4,
         ct_kq1_cond_parq5 = count * kq1_cond_parq5)
```

```{r}
upper_inq5 <- sum(chettey2$ct_kq5_cond_parq5, na.rm = TRUE)
upper_notq5 <- sum(chettey2$ct_kq1_cond_parq5 + chettey2$ct_kq2_cond_parq5 + chettey2$ct_kq3_cond_parq5+ chettey2$ct_kq4_cond_parq5, na.rm = TRUE)

upper_prop <- upper_inq5 / (upper_inq5+upper_notq5)

lower_inq5 <- sum(chettey2$ct_kq5_cond_parq1, na.rm = TRUE)
lower_notq5 <- sum(chettey2$ct_kq1_cond_parq1 + chettey2$ct_kq2_cond_parq1 + chettey2$ct_kq3_cond_parq1+ chettey2$ct_kq4_cond_parq1, na.rm = TRUE)
```

```{r}
lower <- c(lower_inq5, lower_notq5)
upper <- c(upper_inq5, upper_notq5)
table <- rbind(lower, upper)
prop.test(table)
```



Drop levels to make the data set more manageable
```{r}
chet <- chettey2 %>%
  filter(cohort == 1981 ) %>%
  droplevels()
```



##Creating potential groups of interest

```{r}
downward <- chet %>%
  filter(kq1_cond_parq5 > kq5_cond_parq5)
```

More than 30% of student body comes from bottom income quintile
```{r}
down2 <- chet %>%
  filter(par_q1 > .3)
```

```{r}
downward3 <- chet %>%
  filter(kq1_cond_parq5 + kq2_cond_parq5 + kq3_cond_parq5 > kq4_cond_parq5 + kq5_cond_parq5)
```


```{r}
mean(downward3$k_rank)
mean(chet$k_rank, na.rm = TRUE)
```

relative risk aversion



I wanna tell a story about how schools that allow a high percentage of their student body is made up of students from the top quintile it creates an environment that perpetuates the top school

Compare list of colleges with high percentage of parents from top quintile vs those that send a high percentage to the upper class

```{r}
high_incoming_student_top_quintile_schools <- chet %>%
  filter(par_q5 > .5)

high_outgoing_student_top_quintile_schools <- chet %>%
  filter(k_q5 > .5)

A <- high_incoming_student_top_quintile_schools$name
B <- high_outgoing_student_top_quintile_schools$name

in_and_out <- intersect(A, B)

in_not_out <- setdiff(A, B)

out_not_in <- setdiff(B,A)

out_n_in <- chet %>%
  filter(name %in% out_not_in)

in_n_out <- chet %>%
  filter(name %in% in_not_out)

in_a_out <- chet %>%
  filter(name %in% in_and_out) 

in_ <- chet %>%
  filter(name %in% A) %>%
  filter(k_q1 > .2)

```

```{r}
summary(chet$par_q5)
summary(chet$k_q5)


mean(chet$par_q5 , na.rm = TRUE)
median(chet$par_q5 , na.rm = TRUE)
mean(chet$k_q5 , na.rm =  TRUE)
median(chet$k_q5 , na.rm = TRUE)
```




I need to change to categorical with gather!

Then regress k_rank, kq5_cond_parq5, and kq(3 or lower)_cond_parq5
on categorical income. It would look something like:
summary(lm(k_rank ~ par_quintile))

Next Story to investigate
Access rate vs success rate
Potential Graphs to show
plot graph of access rate vs success rate 
  addmitance rate vs k_rank
  ***admittance rate isn't in this data set

plot graph of median parent income vs success rate
  par_median vs k_rank



Graphs I could want

line graph of percentage in kq5 (colored by pq level) to see the change over time
  -this might not be that exciting but I believe it would show
  the unequal distribution of people from the top income is not a new trend
  

I know about the GSS if I find myself looking for new variables to explain something.




